\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-05-03 europa.tex
updates

\eu is implemented as a library that encapsulates all elements
described above making them accessible via modeling and Application
Programming Interface (API) layers.  A user will typically use the
modeling facilities to inject a problem domain description and the
API to extract and manipulate plans and to obtain information about
plan flaws and constraint violations at any stage of the planning
process.

\begin{figure}[b]
\centering
\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
\caption{\small The \eu software architecture.\comment{Is there a
    better block diagram we can use to respond to review comment on
    page 2?}}
\label{fig:europa-architecture}
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships. The \textbf{Constraint
  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
constraints that define relationships among them. It also provides an
efficient arc consistency mechanism \cite{mackworth77}. The
\texttt{CRE} is designed so that specialized reasoning algorithms for
new constraints can be easily and efficiently plugged in. The
\textbf{Plan Database} (\texttt{PDB}) manages object and token
types. It also maintains the state of partial plans in terms of
object, variable and token instances. This constitutes the data
foundation on which end user applications and automated problem
solvers are built. The \textbf{Rules Engine} (\texttt{RE}) modularly
adds the semantics of domain rules that describe dependencies between
tokens. The \textbf{Solver Module} provides the framework for flaw
detection and resolution that is needed to implement search algorithms
for planning. The \textbf{API Layer} provides access to services from
all the different modules in a programmatic way so that \eu can be
easily embedded into specialized applications.

Describing problem domains and problem instances in terms of objects,
variables and domain rules in a programmatic way is cumbersome to the
extent of being impractical; this motivated the creation of modeling
languages that support those definitions in a much more natural and
compact way. \nd (New Domain Description Language
[pronounced ``noodle'']) \cite{NDDL} is the main modeling language for
\eu and will be described in detail in the Modeling section
below. Another, \texttt{ANML} is a language under development that
provides a more intuitive syntax \cite{smith08}.

@@

\item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
  and Solver modules to provide reasoning and search for metric
  resources.

\item \textit{Chronological Backtracking Solver}: A general purpose
  built-in solver to provide the ability to plan out-of-the-box.
  Problems that require challenging search may require specialized
  solvers to be built as explained in Section \ref{sec:europa:search}.

@@

\kcomment{Before going into the details of each module, a simple
  example illustrates how they interact at a high level:}

@@

\kcomment{The snippet above is a \nd (covered in more detail in
  Section \ref{sec:europa:modeling}) model of a light bulb and a
  switch that controls it, along with a problem instance that requires
  the light bulb to be off by time$=10$, given an initial state with
  the bulb on at time$=0$. The \nd module allows \eu to ingest this
  model and the problem description. Once the \nd statements are
  parsed, different object types, object instances and constraints are
  instantiated inside the \texttt{PDB} and \texttt{CRE} modules where
  they are available for both the user and the solver module to query
  and manipulate programmatically. This instantiation of the initial
  state constitutes the initial plan that \eu holds for this
  particular problem instance.

  The user can then invoke functionality available in the solver
  module to modify or extend this initial plan in a way that
  eliminates any flaws (violated constraints, unsatisfied goals,
  oversubscribed resources). When actions are added to a plan, the
  \texttt{RE} module ensures that domain rules expressed in the model
  are enforced when \eu creates new action or predicate instances. For
  instance, if a \texttt{LightSwitch::turnOn} action is added to the
  plan, the \texttt{RE} ensures that requirements for the existence of
  the conditions and effects of that action (that is, the
  corresponding \texttt{LightBulb} must be off before and on after the
  action) are posted through the \texttt{PDB} and \texttt{CRE}
  modules.  }

@@

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different domains and problems, simply by providing
different models and goals. Consequently the expressiveness of the
language for models, goals and plans is of great importance. For \eue,
\nd allows the user to specify the different components of a problem
domain in a precise and concise manner. Among the main features that
make \nd a powerful tool for describing problem domains are that it is
object oriented. \nd supports classes and inheritance in similar
fashion to popular Object-Oriented languages like Java and C++. It
also supports polymorphism for some of the planning components. Using
object classes and instances is a time-tested approach to naturally
describe problem domains \cite{rumbaugh}. It offers declarative
constructs to define constraints and causality dependencies between
actions and effects and finally it offers procedural constructs to
populate \eus \texttt{PDB} with specific problem instances expressed
in terms of objects, variables, constraints, facts and goals. To
articulate the most important elements of \nde, we show how
the Shopping Agent problem could be modeled:

@@

In the model above one can see how the main elements of a problem domain
can be mapped to \eus representation:

@@

\item[\textbf{Object Types}:] \nd supports an object-oriented
  approach, where common data and behavior can be abstracted as a
  class. In this case, we have the \textit{Product} class represent
  the products that the Shopping Agent needs, \textit{Location}
  represents the places where products can be found,
  \textit{ProductLocation} represents the association arising when a
  product can be purchased at a location. The Shopping Agent itself is
  represented by an \textit{Agent} class, which relies on the
  \textit{AgentLocation} class to keep track of where the agent is at
  any point in time. \nd supports a single rooted class
  hierarchy so that state and behavior can be extended through
  inheritance. We take advantage of that feature by making
  \textit{AgentLocation} inherit from \eus built-in \textit{Timeline}
  class which doesn't allow predicates to overlap in time for the same
  object instance. This feature is used for instance, to ensure that
  an agent can only be at one place at a time in our example.

@@

Domain rules can be defined in \nd in the context of a predicate or an
action. In the example above we can see the main elements of a rule
definition:

@@

With that specification for the Shopping Agent domain, a particular
problem instance could be stated as follows:

@@

This problem instance consists of \textbf{Variables} and
\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
\textbf{Constraints}. In our example we define temporal variables that
are used to specify the planning horizon (\texttt{horizonStart} and
\texttt{horizonEnd}) and objects are used to specify locations,
products, product-location associations and the Shopping Agent that we
want to plan for. Facts are used to specify the current state of the
world, do not need any support and are represented as tokens in the
\texttt{PDB}. In this case we use facts to specify that the Shipping
Agent is initially at home. Goals are also represented as tokens in
the \texttt{PDB}. Different from facts, goals trigger inference in \eu
which determines whether flaws (unresolved subgoals, overloaded
resources, etc.) need to be resolved; if so, search modules can be
invoked to resolve those flaws as part of the deliberation
process. Finally, constraints are restrictions on and between facts
and goals. In the example above, we use temporal constraints to
specify when facts hold and when the goals should be accomplished. In
the model above, \texttt{before} and \texttt{after} are \texttt{Allen
  Algebra} \cite{allen84} operators, which \eu supports to specify
temporal relations between tokens (specifically between their start
and end timepoints). The algebra is particularly effective in dealing
with interval arithmetic; its elements are summarized in
Fig. \ref{fig:allen-algebra}

@@

In \eue, all variables and constraints from a problem domain
description are stored inside the \texttt{CRE}.  The \texttt{CRE}
provides a framework where any number of constraint types can
collaborate to perform inference and prune variable domains using
Bounds Propagation and Domain Reduction techniques \cite{marriott98}. 
\eu uses a Propagator pattern which keeps track of a set of
constraint instances of the same type and is responsible for ensuring
variables are updated through inference whenever a relevant change occurs
 (variable domain modifications or constraint addition/deletion).

@@

An \eu user can introduce new constraint types by implementing
specialized constraint and propagator classes; however, \eu provides a
built-in constraint library with a range of arithmetic, temporal, set
and resource constraint types so that a large number of problem
domains can be modeled without having to introduce new constraint
types. Tables \ref{tab:calconst}, \ref{tab:setconst},
\ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
types included in \eus Constraint Library.  Temporal and resource
constraints are especially important to model real-world planning
domains.

\eu uses variables to explicitly represent timepoints for plan
actions and states. Constraints among timepoints provide a natural
way to express domain axioms. For example, in order to state that
activity $A$ must occur before activity $B$ we can say that the end
timepoint of $A$ is $leq$ the start timepoint of $B$.
\cite{dechter91} proposed that constraints among timepoints can be
grouped together to form a Simple Temporal Network
(\texttt{STN}). Such a network can be transformed into a Distance
Graph (\texttt{DG}) where the outward arc from a node represents the
maximum distance from the source node to the target node.
Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
a single constraint along with the resulting \texttt{DG}.
\cite{dechter91} also showed that shortest path algorithms could be
used to propagate values in the network and discover a negative cycle:
namely a path from a node to itself that has a path length less than
$0$. If such a cycle exists, the network is inconsistent. It was
further shown that a single-source shortest path algorithm was
sufficient to detect a negative cycle and provide sufficient
propagation to yield a backtrack-free search. \eus propagator for
temporal constraints is a straight implementation of this algorithm.

\begin{figure}[!htb]
  \centering
  \subfloat[\small A Simple Temporal Network (STN). \kcomment{Nodes correspond
  to variables and their domains and arc's the \emph{constraint} or
  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
  \caption{\small An illustration of a STN and its corresponding
    distance graph.}
  \label{fig:stn}
\end{figure}

Most real-world planning problems need to deal with plans made up of
actions that typically depend upon limited resources (\eg fuel,
storage capacity, battery state of charge, data buffers etc).  % \eu
% provides built-in support for resource computation.
In \eus representation, resource object types come in three forms with
related constraints. \textbf{Reservoir} resources can have both
\texttt{consume(resource,time,quantity)} and
\texttt{produce(resource,time,quantity)} constraints; production and
consumption is instantaneous. \textbf{Reusable} resources on the other
hand, have constraints that use a quantity of the resource for a
specific duration (\texttt{uses(resource,start,duration,quantity)}),
which are then subsequently released. \textbf {Unary} resources are
reusables with capacity$ = 1$, which allows \eu to perform run time
savings in comparison to the generic reusable class. Configuration
options are offered such that in addition to the usual constraint
violations when a resource is completely depleted, plan flaws can be
generated whenever a resource level is outside a specific upper or
lower bound, or when instantaneous production or consumption exceeds a
specific threshold.

Since \eu supports a flexible temporal representation, where time is
represented by intervals, instead of singletons, determining whether a
resource is the source of a plan flaw or a constraint violation in
general is substantially more difficult than just computing a profile
of the resource and checking against some limits.  With temporal
flexibility, a number of resource profiles could be applicable
depending on precisely when each production/consumption event occurs
out of a range of possible values that are in the domain of the
variables that represent the temporal extent for the event.
\cite{Muscettola04,Muscettola06} proposed an algorithm where the
variables that represent the time for production/consumption events
are placed in a temporal network, similar to the network used by the
temporal propagator; constraints on the resource can then be seen as
quantities flowing through that network. Computing the maximum flow
then, yields the maximum resource usage and thus a lower bound on the
resource level; computing a minimum flow will similarly yield an upper
bound. Since maximum flow on a temporal network can be computed in
polynomial time \cite{corman}, \eu takes advantage of this algorithm
to efficiently compute upper and lower bound envelopes on the resource
level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
computation.

\begin{figure}
\centering
\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
\caption{\small An example of resource envelopes computed by \eu.}
\label{fig:resenvelopes}
\end{figure}

Resource envelopes can be useful when trying to evaluate plan
brittleness in the face of changing activity durations or start
times. They can also be useful when searching for plans; however they
are relatively expensive to compute. Therefore their ability to prune
the search space must be weighted against the ability to search faster
\cite{6007772, morris11}.

@@

Since temporal intervals may extend indefinitely in both positive and
negative directions, for problems with a temporal dimension it is also
common to specify a \emph{planning horizon} when invoking a solver;
any decisions that fall completely outside of the horizon can be
ignored. This is useful to ignore activities that may occur far in the
future or in the past to be relevant for a particular plan, ensuring
that the solver is making as few commitments as necessary and reducing
the complexity of plan generation.  The \emph{planning horizon} is
also useful when dealing with recurrent activities; consider for
instance a model where a medical checkup has to occur every $3$
months. The user can model a recurring activity and then use the
horizon to ensure that a finite number of medical checkup activities
are generated as part of a plan.

@@

The initial partial plan is the state of \eus Plan Database after the
initial state has been instantiated. This results in a set of
variable, object and token instances. Inference then takes place to
detect flaws in the partial plan; the three kinds of flaws that can be
detected are:

@@

\item[\textbf{Open Condition}:] an open condition is an
  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
  by explicitly posted goals, or when a token is activated, rules may
  cause the creation of \texttt{INACTIVE} slave tokens as described
  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
  three ways:

@@

  \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
    matching \texttt{ACTIVE} token already in the plan. A token $a$ is said to
    match another token $a?$ iff $a$ and $a?$ unify \cite{russelnorvig}
    and the temporal constraints involving $a$ are satisfied by
    $a?$. Thus, $a$ and $a?$ can be considered the same token
    without the need to introduce $a$ in the plan. Consequently, the
    domain rules associated with $a$ are not introduced into the
    planning process, since they are already triggered when $a?$ was
    introduced in the plan.

  \item[\textbf{Activation}:] A token $a$ is introduced in the
    current plan associating it with the appropriate timeline, without
    choosing a specific time slot; if its timing causes a problem a
    Threat flaw will be triggered (see below).  The domain rules
    associated with $a$ are applied and the subgoal activities
    resulting from those domain rules are introduced as \texttt{INACTIVE}
    tokens.  This results in a number of open condition flaws,
    corresponding to the new subgoal activities.

@@

Open conditions and threats allow flaw detection and resolution at a
higher-level of abstraction (\ie in terms of objects and tokens) than
that of binding variables as is common in Constraint Satisfaction
Problems. This is advantageous when a solver applies heuristics for
ordering choices since it provides a richer context in which to make
decisions.  Furthermore, it aids in reducing the amount of work done
by a solver so that only the necessary refinements are made leaving a
partial plan with sufficient flexibility. For example, one can bypass
committing unbound timepoint variables since threats will force a
solver to impose ordering restrictions on these variables.  Thus the
planning process may yield partially-ordered plans for which all
possible extensions are provably valid.

The Plan Space Planning algorithm described above can be implemented
in multiple ways depending on the approach chosen for flaw and
resolver selection and for backtracking. \eus built-in solver
implements a chronological backtracking algorithm that is summarized
in Algorithm \ref{alg:europa:solve}.

\begin{algorithm}[H]
\KwIn{PartialPlan $plan$}
\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
\Begin{
  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
  \BlankLine
  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
  \BlankLine
  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \comment{\{\emph{potential branching point(s)}\}}\;\label{li:decisionpt}
  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
    \lElse{ $decision.undo()$ \comment{\{\emph{possible backtracking point}\}}\; } \label{li:backtrack}
   }
   \Return \texttt{false} \; \label{li:noplan}
 }
\caption{$\mathrm{bool} ~ solve(plan)$\kcomment{: To deliberate and
    generate partial plans}}
\label{alg:europa:solve}
\end{algorithm}

@@

\item[Line \textbf{\ref{li:backtrack}}:] Otherwise, retract the last
  refinement operation and move on to try the next one.

@@

Algorithm \ref{alg:europa:solve} provides for a sound and complete
search, assuming that no flaws or available refinement operators are
pruned unnecessarily.  Dead-ends in the search are discovered through
constraint propagation, a vehicle for evaluating the consistency of a
partial plan and also for filtering infeasible values from
consideration prior to commitment. Consistency testing is initiated by
the \texttt{isInconsistent} procedure.

The algorithm permits a heuristically controlled search by
applying orderings for \texttt{chooseFlaw} and
\texttt{makeDecisionPoint} resulting in a chronologically-backtracked
search. Heuristics can be implemented by applying ordering/filtering
policies for \texttt{chooseFlaw} and \texttt{makeDecisionPoint}, for
instance, by dealing with earliest flaws first, or by making planning
decisions (open condition flaws) before scheduling decision (threats).

It should be emphasized that while this algorithm is commonly employed
by applications that embed  \eu, it is only one of many that could be
implemented. As a simple example, consider the following solver for
the $N$-Queens problem, where an integer variable V$_i$ represents the
queen in column $i$; V$_1$ $= 3$ implies that the variable on column
$1$ is placed on row $3$.  Non-attack constraints are posted as
arithmetic constraints among the V$_i$ variables to ensure that all
the rows and diagonals contain only one queen.

>>>>

\eu is implemented as a library that encapsulates all elements
\kcomment{previously described} and available via modeling and
Application Programming Interfaces (APIs).  A user will typically use
modeling facilities to inject a problem domain description and the API
to extract and manipulate plans and to obtain information about plan
flaws and constraint violations at any stage of the planning process.

\begin{figure}[b]
\centering
\includegraphics[scale=0.5]{figs/EUROPA-Architecture.pdf}
\caption{\small The \eu software architecture.}
\label{fig:europa-architecture}
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships. The \textbf{Constraint
  Reasoning Engine} (\texttt{CRE}) manages domain types, variables and
constraints that define relationships among them. It also provides an
efficient arc consistency mechanism \cite{mackworth77} designed so
that specialized reasoning algorithms for new constraints can be
easily and efficiently plugged in. The \textbf{Plan Database}
(\texttt{PDB}) manages object and token types. It also maintains the
state of partial plans in terms of object, variable and token
instances. This constitutes the data foundation on which end user
applications and automated problem solvers are built. The
\textbf{Rules Engine} (\texttt{RE}) adds the semantics of domain rules
that describe dependencies between tokens. The \textbf{Solver Module}
provides the framework for flaw detection and resolution that is
needed to implement search algorithms for planning. The \textbf{API
  Layer} provides access to services from all the different modules in
a programmatic way so that \eu can be easily embedded into specialized
applications.

Describing problem domains and problem instances in terms of objects,
variables and domain rules is cumbersome to the extent of being
impractical; this motivated the creation of modeling languages that
support those definitions in a much more natural and compact way. \nd
(New Domain Description Language [pronounced ``noodle'']) \cite{NDDL}
is the \kcomment{primary} modeling language for \eu described in
Section \ref{sec:europa:modeling}. Another, \texttt{ANML}
\cite{smith08} is a language under development that provides a more
intuitive syntax \kcomment{and out of scope for this chapter}.

@@

\item \textit{Resources}: Extensions to the \texttt{CRE}, \texttt{PDB}
  and \textbf{Solver Module} to provide reasoning and search for
  metric resources.

\item \textit{Chronological Backtracking Solver}: A general purpose
  built-in solver to provide the ability to plan out-of-the-box.
  Problems that require challenging search may require specialized
  solvers to be built as noted in Section \ref{sec:europa:search}.

@@

\kcomment{Before going into the details of each module, a simple
  example illustrates how they interact at an abstract level:}

@@

\kcomment{The example above is a \nd (covered in more detail in
  Section \ref{sec:europa:modeling}) model of a light bulb and a
  switch that controls it along with a problem instance that requires
  the light bulb to be \texttt{off} by time$=10$, given an initial
  state with the bulb \texttt{on} at time$=0$. The \nd module allows
  \eu to ingest this model and the problem description. Once the \nd
  statements are parsed, different object types, object instances and
  constraints are instantiated inside the \texttt{PDB} and
  \texttt{CRE} modules where they are available for the user and the
  solver modules to query and manipulate programmatically. This
  instantiation of the initial state constitutes the initial plan that
  \eu holds for this particular problem instance.

  The user can then invoke functionality available in the solver
  module to modify or extend this initial plan in a way that
  eliminates any flaws (violated constraints, unsatisfied goals,
  oversubscribed resources). When actions are added to a plan, the
  \texttt{RE} ensures that domain rules expressed in the model are
  enforced when \eu creates new action or predicate instances. For
  instance, if a \texttt{LightSwitch::turnOn} action is added to the
  plan, the \texttt{RE} ensures that requirements for the existence of
  the conditions and effects of that action (that is, the
  corresponding \texttt{LightBulb} must be off before and on after the
  action) are posted through the \texttt{PDB} and \texttt{CRE}
  modules.  }

@@

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different domains and problems, simply by providing
different models and goals. Consequently the expressiveness of the
language for models, goals and plans is of great importance. For \eue,
\nd allows the user to specify the different components of a problem
domain in a precise and concise manner. Among the main features that
make \nd a powerful tool for describing problem domains are that it is
object oriented. It supports classes and inheritance in similar
fashion to popular object-oriented languages like Java and C++. It
also supports polymorphism for some planning components. Using object
classes and instances is a time-tested approach to naturally describe
problem domains \cite{rumbaugh}. It offers declarative constructs to
define constraints and causality dependencies between actions and
effects and it offers procedural constructs to populate \eus
\texttt{PDB} with specific problem instances expressed in terms of
objects, variables, constraints, facts and goals.

To articulate the most important elements of \nde, we show how the
Shopping Agent problem could be modeled:

@@

The model above the main elements of a problem domain can be mapped to
\eus representation in the \kcomment{following manner}:

@@

\item[\textbf{Object Types}:] \nd supports an object-oriented
  approach, where common data and behavior can be abstracted as a
  class. In this case, we have the \textit{Product} class represent
  the products that the Shopping Agent needs, \textit{Location}
  represents the places where products can be found,
  \textit{ProductLocation} represents the association arising when a
  product can be purchased at a location. The Shopping Agent itself is
  represented by an \textit{Agent} class, which relies on the
  \textit{AgentLocation} class to keep track of where the agent is at
  any point in time. \nd supports a single rooted class hierarchy so
  that state and behavior can be extended through inheritance. We take
  advantage of that feature by making \textit{AgentLocation} inherit
  from \eus built-in \textit{Timeline} class which does not allow
  predicates to overlap in time for the same object instance. This
  feature is used for instance, to ensure that an agent can only be at
  one place at a time in our example.

@@

Domain rules can be defined in \nd in the context of a predicate or an
action. In the example above, the main elements of a rule definition
\kcomment{are as follows}:

@@

With the specification for the Shopping Agent domain \kcomment{above},
a particular problem instance could be stated as:

@@

This problem instance consists of \textbf{Variables} and
\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
\textbf{Constraints}. In our example we define temporal variables that
are used to specify the planning horizon (\texttt{horizonStart} and
\texttt{horizonEnd}) and objects are used to specify locations,
products, product-location associations and the Shopping Agent that we
want to plan for. Facts, represented as tokens in the \texttt{PDB},
are used to specify the current state of the world \kcomment{and} do
not need any support. In this \kcomment{example} we use facts to
specify that the Shipping Agent is initially at home. Goals are also
represented as tokens in the \texttt{PDB}. Different from facts, goals
trigger inference in \eu which determine whether flaws (unresolved
subgoals, overloaded resources, etc.) need to be resolved; if so,
search modules can be invoked to resolve those flaws as part of the
deliberation process. Finally, constraints are restrictions on and
between facts and goals. In the example above, we use temporal
constraints to specify when facts hold and when goals need to be
accomplished. \texttt{before} and \texttt{after} are \texttt{Allen
  Algebra} \cite{allen84} operators, which \eu supports to specify
temporal relations between tokens (specifically between their start
and end timepoints). The algebra is particularly effective in dealing
with interval arithmetic; its elements are summarized in
Fig. \ref{fig:allen-algebra}

@@

In \eue, all variables and constraints from a problem domain
description are stored inside the \texttt{CRE}.  It provides a
framework where any number of constraint types can collaborate to
perform inference and prune variable domains using Bounds Propagation
and Domain Reduction techniques \cite{marriott98}.  \eu uses a
Propagator pattern which keeps track of a set of constraint instances
of the same type and is responsible for ensuring variables are updated
through inference whenever a relevant change occurs (variable domain
modifications or constraint addition/deletion).

@@

An \eu user can introduce new constraint types by implementing
specialized constraint and propagator classes. However \eu provides a
built-in constraint library with a range of arithmetic, temporal, set
and resource constraint types so that a large number of problem
domains can be modeled. Tables \ref{tab:calconst}, \ref{tab:setconst},
\ref{tab:objhierarchy} and \ref{tab:misc} show the main constraint
types included in \eus Constraint Library.  Temporal and resource
constraints are especially important to model real-world planning
domains.

\eu uses variables to explicitly represent timepoints for plan actions
and states. Constraints among timepoints provide a natural way to
express domain axioms. For example, in order to state that activity
$A$ must occur before activity $B$ we can say that the end timepoint
of $A$ is $leq$ the start timepoint of $B$.  \cite{dechter91} proposed
that constraints among timepoints can be grouped together to form a
Simple Temporal Network (\texttt{STN}). Such a network can be
transformed into a Distance Graph (\texttt{DG}) where the outward arc
from a node represents the maximum distance from the source node to
the target node.  \kcomment{They} also show that shortest path
algorithms \kcomment{(see \cite{corman} for a review)} could be used
to propagate values in the network and discover a negative cycle:
namely a path from a node to itself that has a path length less than
$0$. If such a cycle exists, the network is \kcomment{provably}
inconsistent. It was further shown that a single-source shortest path
algorithm was sufficient to detect a negative cycle and provide
sufficient propagation to yield a backtrack-free search. \eus
propagator for temporal constraints is an implementation of such a
single-source algorithm. Fig. \ref{fig:stn} illustrates an
\texttt{STN} with $2$ variables and a single constraint along with the
resulting \texttt{DG}.

\begin{figure}[!htb]
  \centering
  \subfloat[\small A Simple Temporal Network (STN). \kcomment{Nodes correspond
  to variables and their domains and arc's the \emph{constraint} or
  distance between them.}]{\label{fig:stn1}\includegraphics[width=0.4\columnwidth]{figs/stn-graph.pdf}}
  \subfloat[\small The resulting distance graph from \ref{fig:stn1}.]{\label{fig:stn2}\includegraphics[width=0.4\columnwidth]{figs/distance-graph.pdf}}
  \caption{\small An illustration of a Simple Temporal Network
    (\texttt{STN}) and its corresponding distance graph.}
  \label{fig:stn}
\end{figure}

Most real-world planning problems need to deal with plans made up of
actions that typically depend upon limited resources (\eg fuel,
storage capacity, battery state of charge, data buffers etc).  % \eu
% provides built-in support for resource computation.
In \eus representation, resource object types come in three forms with
related constraints. \textbf{Reservoir} resources can have both
\texttt{consume(resource,time,quantity)} and
\texttt{produce(resource,time,quantity)} constraints; production and
consumption is instantaneous. \textbf{Reusable} resources on the other
hand, have constraints that use a quantity of the resource for a
specific duration (\texttt{uses(resource,start,duration,quantity)}),
which are then subsequently released. \textbf {Unary} resources are
reusables with capacity$=1$, which \kcomment{accounts for} runtime
savings in comparison to the generic reusable class. Configuration
options are offered such that in addition to the usual constraint
violations when a resource is completely depleted, plan flaws can be
generated whenever a resource level is outside a specific bound, or
when instantaneous production or consumption exceeds a specific
threshold.

Determining whether a resource is the source of a plan flaw or a
constraint violation in general is substantially more difficult in
\eus flexible representation, than just computing a profile of the
resource and checking against some limits.  With temporal flexibility,
a number of resource profiles could be applicable depending on
precisely when each production/consumption event occurs out of a range
of possible values that are in the domain of the variables that
represent the temporal extent for the event.
\cite{Muscettola04,Muscettola06} propose an algorithm where the
variables that represent the time for production/consumption events
are placed in a temporal network, similar to the network used by the
temporal propagator; constraints on the resource can then be seen as
quantities \emph{flowing} through that network. Computing the maximum
flow then, yields the maximum resource usage and thus a lower bound on
the resource level; computing a minimum flow will similarly yield an
upper bound. They show that such upper and lower bound envelopes on
the resource level can be computed in polynomial time
\cite{corman}. Resource envelopes can be useful when trying to
evaluate plan brittleness in the face of changing activity durations
or start times. They can also be useful when searching for plans;
however they are relatively expensive to compute. Therefore their
ability to prune the search space must be weighted against the ability
to search faster \cite{6007772, morris11}. Fig. \ref{fig:resenvelopes}
shows an example of such a resource computation.

\begin{figure}
\centering
\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
\caption{\small An example of resource envelopes \cite{Muscettola04,Muscettola06} computed by \eu.}
\label{fig:resenvelopes}
\end{figure}

@@

Since temporal intervals may extend indefinitely in both positive and
negative directions \kcomment{(\eg $[-\infty,+\infty]$)}, for problems
with a temporal dimension it is also common to specify a
\emph{planning horizon} when invoking a solver. Any decisions that
fall completely outside of the horizon can be ignored, ensuring that
the solver is making as few commitments as necessary and reducing plan
generation complexity.  The \emph{planning horizon} is also useful
when dealing with recurrent activities. Consider for instance a model
where a medical checkup has to occur every $3$ months. The user can
model a recurring activity and then use the horizon to ensure that a
finite number of medical checkup activities are generated as part of a
plan.

@@

The initial partial plan is the state of \eus \texttt{PDB} after the
initial state has been instantiated. This results in a set of
variable, object and token instances. Inference then takes place to
detect flaws in the partial plan. The three kinds of flaws that can be
detected are:

@@

\item[\textbf{Open Condition}:] an open condition is an
  \texttt{INACTIVE} token which can be generated by explicitly posted
  goals or during token activation when rules may cause the creation
  of \texttt{INACTIVE} slave tokens as described in Section
  \ref{sec:europa:pr}. Open Conditions can be resolved in three ways:

@@

  \item[\textbf{Merging}:] An \texttt{INACTIVE} token is merged with a
    matching \texttt{ACTIVE} token already in the plan. A token $a$ is
    said to match another token $a?$ iff $a$ and $a?$ unify
    \cite{russelnorvig} and the temporal constraints involving $a$ are
    satisfied by $a?$. Thus, $a$ and $a?$ can be considered the same
    token without the need to introduce $a$ in the plan. Consequently,
    the domain rules associated with $a$ are not introduced into the
    planning process, since they are already triggered when $a?$ was
    \kcomment{instantiated} in the plan.

  \item[\textbf{Activation}:] A token $a$ is introduced in the current
    plan associating it with the appropriate timeline, without
    choosing a specific time slot; if its timing causes a problem a
    Threat flaw will be triggered (see below).  The domain rules
    associated with $a$ are applied and the subgoal activities
    resulting from those domain rules are introduced as
    \texttt{INACTIVE} tokens.  This results in a number of open
    condition flaws corresponding to the new subgoal activities.

@@

Open conditions and threats allow flaw detection and resolution at a
higher-level of abstraction (\ie in terms of objects and tokens) than
that of binding variables as is common in Constraint Satisfaction
Problems. This is advantageous when a solver applies heuristics for
ordering choices since it provides a richer context in which to make
decisions.  Furthermore, it aids in reducing the amount of work done
by a solver so that only the necessary refinements are made,
\kcomment{ensuring} a partial plan with sufficient flexibility. For
example, one can bypass committing unbound timepoint variables since
threats will force a solver to impose ordering restrictions on these
variables.  Thus the planning process may yield partially-ordered
plans for which all possible extensions are provably valid.

The Plan Space Planning algorithm described can be implemented in
multiple ways depending on the approach chosen for flaw and resolver
selection and for backtracking. \eus built-in solver implements a
chronological backtracking algorithm that is summarized in Algorithm
\ref{alg:europa:solve}.

\begin{algorithm}
\KwIn{PartialPlan $plan$}
\KwOut{\texttt{true} if a plan solution was found, \texttt{false} otherwise}
\Begin{
  \lIf{isInconsistent(plan)}{\Return \texttt{false}\;}\label{li:propagate}
  \BlankLine
  Flaw $flaw \leftarrow chooseFlaw(plan)$ \; \label{li:selectflaw}
  \lIf{ $\emptyset = flaw$ }{\Return \texttt{true} \;} \label{li:noflaw}
  \BlankLine
  DecisionPoint $decision \leftarrow makeDecisionPoint(flaw, plan)$ \comment{\{\emph{potential branching point(s)}\}}\;\label{li:decisionpt}
  \While{ $decision.hasNext()$ }{ \label{li:decisionloop}
    PartialPlan $pp \leftarrow decision.executeNext()$ \; \label{li:decide}
    \lIf{ $solve(pp)$ }{ \Return \texttt{true} \; } \label{li:recurse}
    \lElse{ $decision.undo()$ \comment{\{\emph{possible backtracking point}\}}\; } \label{li:backtrack}
   }
   \Return \texttt{false} \; \label{li:noplan}
 }
\caption{$\mathrm{bool} ~ solve(plan)$\kcomment{: To deliberate and
    generate partial plans}}
\label{alg:europa:solve}
\end{algorithm}

@@

\item[Line \textbf{\ref{li:backtrack}}:] Otherwise, retract the last
  refinement operation and move on to try the next.

@@

Algorithm \ref{alg:europa:solve} provides for a sound and complete
search, assuming that no flaws or available refinement operators are
pruned unnecessarily.  Dead-ends in the search are discovered through
constraint propagation, a vehicle for evaluating the consistency of a
partial plan and also for filtering infeasible values from
consideration prior to commitment. Consistency testing is initiated by
the \texttt{isInconsistent} procedure.  The algorithm permits a
heuristically controlled search by applying orderings for
\texttt{chooseFlaw} and \texttt{makeDecisionPoint} resulting in a
chronologically-backtracked search. Heuristics can be implemented by
applying ordering/filtering policies for \texttt{chooseFlaw} and
\texttt{makeDecisionPoint}, for instance, by dealing with earliest
flaws first, or by making planning decisions (open condition flaws)
before scheduling decision (threats).

It should be emphasized that while this algorithm is commonly employed
by applications that embed \eue, it is only one of many that could be
implemented. As an example, consider the solver in Algorithm
\ref{algo:n-queens} for the $N$-Queens problem, where an integer
variable V$_i$ represents the queen in column $i$; V$_1$ $= 3$ implies
that the variable on column $1$ is placed on row $3$.  Non-attack
constraints are posted as arithmetic constraints among the V$_i$
variables to ensure that all the rows and diagonals contain only one
queen. The \texttt{CRE} can be used to compute the non-attack
constraint violations, with \texttt{ce.getViolations()} a call on the
\eu API. The \texttt{CRE} can also be used to pick the queen that has
the most violations in \texttt{getQueenWithMaxViolation()} and to look
ahead and see what moves are possible for a queen that would remove
its violations in \texttt{getMoves()}.
